---
title: "SEAWINDS"
author: "Simon Lacombe"
date: "03/03/2021"
output: html_document
---

<style>
body {
text-align: justify}
</style>


```{r setup, include=FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE, results = "hide"}
sapply(c( 'raster','stringr', 'data.table', 'magrittr', 'sf', "rnaturalearthdata","rnaturalearth", 'windR', 'geosphere', 'ggplot2'),
       function(x) suppressPackageStartupMessages(require(x , character.only = TRUE, quietly = TRUE)))

pxdata = readRDS("outputs/monthly_pixel_data.rds")
cosd <-function(x) cos(x*pi/180) 
sind <-function(x) sin(x*pi/180)
```

# 1.Background

Birds are among the animals that exhibit the longest migratory journeys, which can span up to 50 000 km every year (Hromádková et al., 2020). Migratory birds generally spend summer at higher latitudes in colonies where they gather in large numbers to breed. In autumn, they move to more productive regions to winter in . Consequently, they undertake every year a round trip migration between their wintering range and their breeding range (Newton, 2010). In addition, within a given season, birds may also cover relatively long distances. For instance, birds can display nomadic movements between different foraging areas in winter **[Ref]** or commute between the colony and a foraging area during breeding, when feeding their chick (Tarroux et al., 2016). Overall, birds can fly very long distances within a year.  

Flying over such long distances involves important energy expenditures (Ward, 2004). Thus, birds display many physiological and behavioral adaptations to reduce the energy costs associated with flying (Videler, 2006). One way to reduce theses costs is to take advantage of the wind conditions. For instance, a bird flying tailwind will spend less energy and go faster than a bird flying with no wind or with headwind (Elliott et al., 2014; Krietsch et al., 2020; Safi et al., 2013). Furthermore, although birds seem to be able to compensate for the drift associated with crosswind by shifting their direction, this compensation is costly and reducing the amount of crosswind encountered during the flight would reduce the time spent flying and the energy expenditure (Newton, 2010; Tarroux et al., 2016).

Many arctic bird species have thus been shown to adjust their flights according to the prevailing winds. For example, during their migration between the Arctic and Antarctica, Arctic terns *Sterna paradisaea* adjust their route to benefit from tailwind support, especially during the northbound migration before breeding (Hromádková et al., 2020).  Pectoral sandpiper *Calidris melanotos* also adjust their movement to the prevailing winds. During the breeding season, male pectoral sandpipers sample several colonies to maximize their breeding success.  Because mating opportunities in the colonies are unpredictable, the site they choose to visit is greatly influenced by the wind conditions at departure (Krietsch et al., 2020). Another interesting example is the one of Antarctic petrels, *Thalassoica antarctica*. These birds fly over long distances between their colony and their foraging areas with very strong winds. They have been shown to be able to adjust their flight altitude to minimize the effect of wind and to adjust their heading to compensate for the drift (Tarroux et al., 2016).

The way wind influences flight patterns greatly depends on the species’ behavior and the type of flight considered. Some birds are highly philopatric and will always return to the same breeding and wintering areas, using the same migratory routes **[Ref]**. Other birds, like the pectoral sandpiper, are more flexible and show little to no site-fidelity (Krietsch et al., 2020). We can expect that philopatric species will be more constrained in their migratory routes and will adjust it to take advantage of winds less often. The target site predictability may also be involved in determining the tendency of the bird to use wind, with birds flying towards unpredictable destinations using it more often. Whether philopatry and site predictability effectively influences the birds’ ability to use wind while migrating remains unknown. 

Migratory routes and timings are also constrained by several factors other than wind among which phenology plays an important part. Indeed, when birds migrate towards their breeding site, arriving early can greatly improve their mating success **[Ref]**. Thus, we can expect that birds with an important constraint on breeding timing will be less likely to wait for favorable winds at departure (Note that the timing of migration can be impacted by factors other than breeding. For instance, diurnal foragers could be advantaged when crossing the equator near the equinox to experience the most daylight (Hromádková et al., 2020)). During their migrations, stopovers are also important for birds to restore energy. Foraging opportunities along the migratory track are thus an important determinant of the migratory route (Alerstam, 2011). For instance, foraging opportunities at stopovers have been shown to strongly impact the migratory route of Arctic terns, especially during their post breeding migration (Hromádková et al., 2020). For Cory’s shearwater *Calonectris borealis*, most individuals fly with the wind towards their wintering area in the south of the Atlantic Ocean. But several individuals fly North with headwind, probably to take advantage of reduced competition along the way and in the wintering site according to the ideal free distribution theory (Dell’Ariccia et al., 2018). 

The black-legged kittiwake, *Rissa tridactyla*, is a pelagic seabird breeding at high latitudes along the Northern coasts of the Atlantic and Pacific (Daunt et al., 2002; Hatch et al., 2020). In winter, they migrate to the open ocean, with their non-breeding range covering a great part of both the Atlantic and the Pacific Ocean (Hatch et al., 2020). Their diet is mainly composed of fish, they may also feed on zooplankton (Mehlum and Gabrielsen, 1993). During the breeding period, they act as central place foragers, splitting their time between foraging flights (traveling flight and active foraging flights) and attending at the nest (Daunt et al., 2002). Daunt et al. (2002) showed that during this period, kittiwakes stay in close range to their breeding site (upper limit ca. 73km for kittiwakes on the Isle of May). This upper limit may be the result of both prey availability around the colony and the energetic costs associated with the traveling flight. Whether the wind conditions influence the range of the foraging flights around the colony, and whether it impacts the foraging efficiency during the breeding period is unknown.

Little is known concerning their movements outside the breeding season. Most kittiwakes from European colonies travel across the North-Atlantic to winter in the Labrador sea (Frederiksen et al., 2012; Swindells, n.d.). What determines the wintering? site of a given individual is unknown, with the principal wintering areas being shared between birds from many different colonies, whereas the birds from a given colony can winter in sites several thousand kilometers apart (Frederiksen et al., 2012; Orben et al., 2015; Swindells, n.d.). Kittiwakes generally winter in highly productive areas like oceanic gyres, shelf edges and upwelling zones (Frederiksen et al., 2012; Orben et al., 2015). These areas seem to provide more predictable resources (Weimerskirch, 2007) and kittiwake could thus specifically target them, travelling long distances without clear knowledge of the conditions at their arrival site. Furthermore, although black-legged kittiwakes are highly flexible regarding their wintering areas (important interannual variability), individual experiences seem to have an impact on the wintering destination of the birds, with individuals returning to previously visited areas more often than expected by chance (Orben et al., 2015). Together, prey predictability and individual experiences could explain the difference in wintering areas within and across colonies. Yet, for black-legged kittiwakes flying costs are high (Elliott et al., 2014) and reducing the energy expenditure to get to the foraging area could also impact their wintering distribution. **Wind may thus have an impact on their target destination by reducing the energy cost of the flight**. We could imagine that (1) wind at departure and during the flight partially determines the place of settlement; or that (2) birds time their departure according to the dominant wind to help them going to a determined place. In addition, beyond their migratory journey, foraging flight at their wintering location may represent important distances (no data found) and black-legged kittiwakes could adapt to wind conditions (either by avoiding strong wind, or by adjusting their heading to benefit from tailwind) in order to maximize the energy income while foraging. Finally, the effect of the weather conditions encountered in winter on their breeding success is unknown and it would be interesting to assess it.

# 2. Methods used to study wind and migrations

## Wind data:

***Several Datasets can be used***

* Direct measurements: **ASCAT** (scatterometry). Seems like ASCAT is the only direct measure available covering the period of interest. 

**ASCAT: 1 day; 0.25°; 2013-present; surface **

* Atmospheric reanalysis: **ERA-Interrim, ECMWF**. Reanalysis combine measurements with a forecast model. The fact that it is based on observations (*e.g.* several scatterometers used like QUICKSCAT before 2011) makes it a strong dataset. 
These data are used in (Hromádková et al., 2020; Krietsch et al., 2020) and (Dell’Ariccia et al., 2018) uses another reanalysis from the NOAA.

**ERA-Interrim: 6 hr; 0.75° (interpolated data are available); covering the whole period; several altitudes.**

##	Light logger technology: CF (Lisovski et al., 2020)

## Vectors used 

```{r, echo = FALSE}
i = 3

ggplot(data=pxdata[which(not(is.na(pxdata$gdir)))[i]]) + 
  geom_segment(aes(x = 0, y = 0, xend = gspeed*sind(gdir), yend = gspeed*cosd(gdir)), color = 'red', size = 1, arrow = arrow(length = unit(1, "cm"))) + #Ground 
  geom_segment(aes(x = 0, y = 0, xend = wspeed*sind(wdir), yend = wspeed*cosd(wdir)), color = 'blue', size = 1, arrow = arrow(length = unit(1, "cm"))) + #Wind 
  geom_segment(aes(x = 0, y = 0, xend = ws*sind(gdir), yend = ws*cosd(gdir)), color = 'black', size = 1, arrow = arrow(length = unit(1, "cm"))) + #wind support 
  geom_segment(aes(x = 0, y = 0, xend = cw*sind(gdir+90), yend = cw*cosd(gdir+90)), color = 'black', size = 1, arrow = arrow(length = unit(1, "cm"))) + #crosswind
  geom_segment(aes(x=wspeed*sind(wdir), y =wspeed*cosd(wdir), xend = wspeed*sind(wdir) + aspeed*sind(adir), yend = wspeed*cosd(wdir) + aspeed*cosd(adir)), color = 'green', size = 1, arrow = arrow(length = unit(1, "cm"))) +#air
ggtitle("Vectors used to characterize the bird’s movement in relation with the wind.\n Red: ground vector; Blue: wind vector; green: air vector;\n Black: cross wind and wind support vectors. (Left: WS < 0; Right: WS > 0)")
```

## Biases, spatial and temporal resolution, ... 
CF Safi & al, 2013

# 3. Problematic of the study
Do black legged kittiwakes take advantage of winds (and currents?) when they migrate? Is their route adapted to take advantage of winds?

1. Wind support on the actual track VS on randomly created tracks (we can use IRMA and only specify departure and arrival points (Fauchald et al., 2019)) or VS wind support bird would have encountered flying outbound by the inbound track and conversely (I’m afraid this second method won’t be informative enough) or VS strait line.

2.	I can make a “least cost map” (in terms of energy expenditure i.e. travel time + wind support) between two locations (a breeding colony and a wintering site) and study whether this map predicts the birds’ track. It could provide us with a map predicting the migratory routes based on wind data.

**NB**: I like the idea of a least cost map. We could for instance divide the studied area into a grid (same resolution as the wind data) and plot for each point the travel time (how to create a relevant proxy of the energy expenditure? 1/travel time? Wind support/travel time?) to go from departure to arrival by a bid that would pass by this point.

3. If we have enough time, an aspect I’m very interested in is the interannual (and intercolonial) variability of wind support: for instance, no annual variability in WS as well as a variability in wind conditions and migratory tracks could be indicative of behavioral adjustment to wind conditions (be it a shift in the migratory route or the destination).

**NB**: I’ll focus on the whole track, but it could be interesting/necessary to compare the migratory segments and the stationary ones. I can access the output of Françoise’s Lavielle segmentation to distinguish between the two types of movements. Maybe other types of segmentation could be necessary in our particular case (*e.g* residence time (CF Dell'Arccia & al.))
# 4. Monthly analysis 

I plotted on maps of the North Atlantic Ocean the mean wind direction and speed, and the mean ground direction and speed of the birds (see below). 

```{r, echo = FALSE, warning = FALSE}
world <- ne_countries(scale = "medium", returnclass = "sf")
a=4 #Arrow size

ggplot(data = pxdata[not(is.na(pxdata$n))]) + 
  geom_segment(aes(x = x, y = y, xend = x+gspeed/a*sind(gdir), yend = y+gspeed/a*cosd(gdir), color=ws),arrow = arrow(length = unit(.1, "cm"))) +
  scale_color_gradient("wind support", low = "red", high = "green") +
  geom_sf(data=world ,fill = "black", color = "black") + 
  coord_sf(xlim = c(min(pxdata$x)-1, max(pxdata$x)+1), ylim = c(min(pxdata$y)-1, max(pxdata$y)+1), expand = FALSE)+
  ggtitle("Map of the Atlantic Ocean showing the wind vectors")
ggplot(data = pxdata) +
  geom_segment(aes(x = x, y = y, xend = x+wspeed/a*sind(wdir), yend = y+wspeed/a*cosd(wdir)),arrow = arrow(length = unit(.1, "cm"))) +
  geom_sf(data=world ,fill = "black", color = "black") + 
  coord_sf(xlim = c(min(pxdata$x)-1, max(pxdata$x)+1), ylim = c(min(pxdata$y)-1, max(pxdata$y)+1), expand = FALSE)+
  ggtitle("Map of the Atlantic ocean showing the birds’ movement.\n The wind support encountered by the birds is represented with a gradient color.")


```


***In order to study the relationship between wind and the birds’ movement, I make some predictions that I will test:***

1.	Ground direction on a pixel is correlated with wind direction on the same pixel.

I don't know how to test this correlation. I tried with a circular correlation coefficient and found $r^2 \simeq -0.23$ with a p-value$ < 0.001$. I don't know how relevant this coefficient is. I also found that a coefficient called $\rho_v^2$ would be relevant (many references using this coefficients) but i don't know how to compute it with R. 

2.	Overall, WS > 0. On migratory segments, this effect shall be stronger.

```{r, echo = FALSE, warning = FALSE, message = FALSE}
 ggplot(data = pxdata)+ #histogram of wind support 
    geom_histogram(aes(x=ws))+
    xlab("Wind support")
  print(paste0("mean wind support :", mean(pxdata$ws, na.rm = T)))
```


3.	If the bird maximizes travel speed: Gspeed increases with WS and airspeed is constant. If the birds minimize the energy expenditure: Gspeed is constant but airspeed decreases with WS. We can imagine that birds on migratory segments maximize travel speed while they minimize the energy expenditure on stationary segments.

```{r, echo = FALSE, warning = FALSE, message = FALSE}
ggplot(data = pxdata)+ #Air speed as a function of wind support
    geom_point(aes(x=ws, y = aspeed, size = gspeed))+
    geom_smooth(aes(x=ws, y = aspeed), method = "lm")+
    xlab("Wind support")+ylab("air speed")
  
  ggplot(data = pxdata)+ #Ground speed as a function of wind support
    geom_point(aes(x=ws, y = gspeed))+
    geom_smooth(aes(x=ws, y = gspeed), method = "lm")+
    xlab("Wind support")+ylab("ground speed")
  
```


4.	WS is higher on the migratory path compared with the straight line.  

I can maybe test the hypothesis : Wind Support is statistically higher along the migratory route (*i.e* between Greenland and Iceland) with an ANOVA.

